---
title: "Homework 10"
author: "Ross Brancati"
date: "11/15/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
#before we begin, load required libraries
library(alr4) #Data to accompany textbook
library(tinytex) #Light version of latex
#and set working directory
setwd("/Users/rossbrancati/Documents/Classes/fall_2022/stat_625/homeworks/homework_10")
```

## 1: Problem 11.1
### 11.1.1
This is a nonlinear mean function because the slope parameter $\beta_{ij}$ is multiplied by a parameter $\gamma$. 

Since G has many levels, there is a mean function for each level of G. Additionally, each of the mean functions has its own slope since the slope parameter $\beta_{ij}$ is dependent on the level of G, and the lines are concurrent when x = $\gamma$.

### 11.1.2
First, get the starting values
```{r}
#fit model
m1 <- lm(TS ~ log(BodyWt):factor(D), data=sleep1) 
#summarize
summary(m1)$coef
```

Next, fit the nonlinear model with coefficients found above and gamma (g) = 0
```{r}
#fit model
m2 <- nls(TS ~ 
            b0 + 
            b11*((D==1)*(log(BodyWt) - g)) + 
            b12*((D==2)*(log(BodyWt) - g)) + 
            b13*((D==3)*(log(BodyWt) - g)) + 
            b14*((D==4)*(log(BodyWt) - g)) + 
            b15*((D==5)*(log(BodyWt) - g)),
          data=sleep1,
          start=list(b0=11, 
                     b11=-.3, 
                     b12=-.6, 
                     b13=-.9, 
                     b14=-.6, 
                     b15=-1.6, 
                     g=0))
#summarize
summary(m2)
```

